---
title: "An Introduction to `heemod`"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{An Introduction to heemod}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, include=FALSE}
library(heemod)
```

This document is a presentation of the basic steps to define and run a model in `heemod`. Note that decision trees are actually a subset even of *Markov model*, and thus can be specified easily with this package.

When building a Markov model the following steps must be performed:

  * Specify transition probabilities between states.
  * Specify values attached to states (costs, utilities, etc.).
  * Combine this information and run the model.

## Transition probabilities

The probability to transition from one state to another during a time period is called a *transition probability*. Said time period is called a *cycle*.

Transition probabilities between states can be described through a 2-way table where the lines correspond to the states at the beginning of a cycle and the columns to the states at the end of a cycle. Consider a model with 2 states `A` and `B`:

|       | A | B |
|-------|---|---|
| **A** | 1 | 2 |
| **B** | 3 | 4 |

When starting a cycle in state **A** (row **A**), the probability to still be in state **A** at the end of the cycle is found in colunm **A** (cell **1**) and the probability to change to state **B** is found in column **B** (cell **2**).

Similarly, when starting a cycle from state **B** (row **B**), the probability to be in state **A** or **B** at the end of the cycle are found in cells **3** or **4** respectively.

In the context of Markov models, this 2-way table is called a *transition matrix*. A transition matrix can be defined easily in `heemod` with the `define_transition()` function. If we consider the previous example, where cell values have been replaced by actual probabilities:

|       |  A  |  B  |
|-------|-----|-----|
| **A** | 0.9 | 0.1 |
| **B** | 0.2 | 0.8 |

That transition matrix can be defined with the following command:

```{r}
mat_trans <- define_transition(
  .9, .1,
  .2, .8
)
mat_trans
```

## Attach values to states

Values are attached to states. Cost and utility are classical examples of such values. To continue with the previous example, the following values can be attachd to state **A** and **B**:

  * State **A** has a cost of *1234* per cycle and an utility of *0.85*.
  * State **B** has a cost of *4321* per cycle and an utility of *0.50*.
  
A state and its values can be defined with `define_state()`:

```{r}
state_A <- define_state(
  cost = 1234,
  utility = 0.85
)
state_A

state_B <- define_state(
  cost = 4321,
  utility = 0.50
)
state_B
```

## Combine information in a model

Now that the transition matrix and the state values are defined for a given strategy, we can combine them with `define_strategy()`:

```{r}
strat <- define_strategy(
  transition = mat_trans,
  state_A,
  state_B
)
strat
```

## Run the model

A model is a made of one or more strategy, run with `run_model()` for a given number of cycles. Here we run only one strategy, so no comparison of the cost-effectiveness of the different strategies will be made. The variables corresponding to valuation of cost and effect must be given at that point.

```{r}
res_mod <- run_model(
  strat,
  cycles = 10,
  cost = cost,
  effect = utility
)
res_mod
```

By default the model is run for *1000* persons starting in the first state (here state **A**).

## Analyse results

We can plot the state membership counts over time. Other plot types are available.

```{r, fig.width = 6, fig.height=4, fig.align='center'}
plot(res_mod)
```

Plots can be modified using `ggplot2` syntax.

```{r, fig.align='center', fig.height=4, fig.width=6, message=FALSE}
library(ggplot2)

plot(res_mod) +
  xlab("Time") +
  ylab("N") +
  theme_minimal() +
  scale_color_brewer(
    name = "State",
    palette = "Set1"
  )
```

And black & white plots for publication are available with the `bw` plot option

```{r, fig.align='center', fig.height=4, fig.width=6, message=FALSE}
library(ggplot2)

plot(res_mod, bw = TRUE)
```

The state membership counts and the values can be accessed with `get_counts()` and `get_values()` respectively.

```{r}
head(get_counts(res_mod))
head(get_values(res_mod))
```

## Convenience functions

Convenience functions are available to easily compute transition probabilities from indidence rates, OR, RR, or probabilities estimated on a different timeframe.

Example : convert an incidence rate of *162* cases per *1,000* person-years to a 5-year probability.

```{r}
rate_to_prob(r = 162, per = 1000, to = 5)
```

See `?probability` to see a list of the convenience functions available.

## External data

Mortality rates by age and sex (often used as transition probabilities) can be downloaded from the WHO online database with the function `get_who_mr()`.

External data contained in user-defined data frames can be referenced in a model with the function `look_up()`.

## Compare strategies

In order to compare different strategies it is possible to run a model with multiple strategies in parallel, examples are provided in `vignette("c-homogeneous", "heemod")` or `vignette("d-non-homogeneous", "heemod")`.

## Time-variability

Time varying transition probabilities and state values are available through the `markov_cycle` and `state_cycle` variables. Time-varying probabilities and values are explained in `vignette("b-time-dependency", "heemod")`.

## Transition costs

Transition costs between states can be defined by adding an additional state called a transition state. This is a special case of a tunnel state. A situation where the transtion from `A` to `B` (noted `A->B`) costs \$100 and reduces utility by 0.1 points compared to the usual values of `B` can be modelled by adding a state `B_trans`.

The cost and utility of `B_trans` are the same as those of `B` +\$100 and -0.1 QALY respectively. 
`A->B_trans` is equal to the former value of `A->B`. `A->B` is set to 0 (all transitions from `A` to `B` must pass through `B_trans` from now on). 
The probability to stay in `B_trans` is 0, `B_trans->B` is equal to `B->B` and similarly all `B_trans->*` are equal to `B->*`.

## Parallel computing

Some operation such as PSA can be significantly sped up using parallel computing. This can be done in the following way:

  * Define a cluster with the `use_cluster()` functions (i.e. `use_cluster(4)` to use 4 cores).
  * Run the analysis as usual.
  * To stop using parallel computing use the `close_cluster()` function.
  
  
Results may vary depending on the machine, but we found speed gains to be quite limited beyond 4 cores.

## Going further

Probabilistic uncertainty analysis `vignette("e-probabilistic", "heemod")` and deterministic sensitivity analysis `vignette("f-sensitivity", "heemod")` can be performed. Population-level estimates and heterogeneity can be computed : `vignette("g-heterogeneity", "heemod")`.
