---
title: "Demographic Analysis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Demographic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, include=FALSE}
library(heemod)
```

```{r, include = FALSE}

param <- define_parameters(
    age_init = 60,
    sex = 0,
    # age increases with cycles
    age = age_init + markov_cycle,
    
    # operative mortality rates
    omrPTHR = .02,
    omrRTHR = .02,
    
    # re-revision mortality rate
    rrr = .04,
    
    # parameters for calculating primary revision rate
    cons = -5.49094,
    ageC = -.0367,
    maleC = .768536,
    lambda = exp(cons + ageC * age_init + maleC * sex),
    gamma = 1.45367786,
    
    rrNP1 = .260677,
    
    # revision probability of primary procedure
    standardRR = 1 - exp(lambda * ((markov_cycle - 1) ^ gamma -
                                     markov_cycle ^ gamma)),
    np1RR = 1 - exp(lambda * rrNP1 * ((markov_cycle - 1) ^ gamma - 
                                        markov_cycle ^ gamma)),
    
    # age-related mortality rate
    sex_cat = ifelse(sex == 0, "FMLE", "MLE"),
    mr = get_who_mr(age, sex_cat, country = "GBR"),
    
    # state values
    u_SuccessP = .85,
    u_RevisionTHR = .30,
    u_SuccessR = .75,
    c_RevisionTHR = 5294
)

mat_standard <- define_matrix(
    state_names = c(
      "PrimaryTHR",
      "SuccessP",
      "RevisionTHR",
      "SuccessR",
      "Death"
    ),
    0, C, 0,          0, omrPTHR,
    0, C, standardRR, 0, mr,
    0, 0, 0,          C, omrRTHR+mr,
    0, 0, rrr,        C, mr,
    0, 0, 0,          0, 1
)

mat_np1 <- define_matrix(
    state_names = c(
      "PrimaryTHR",
      "SuccessP",
      "RevisionTHR",
      "SuccessR",
      "Death"
    ),
    0, C, 0,          0, omrPTHR,
    0, C, np1RR,      0, mr,
    0, 0, 0,          C, omrRTHR+mr,
    0, 0, rrr,        C, mr,
    0, 0, 0,          0, 1
)

mod_standard <- define_model(
  transition_matrix = mat_standard,
  PrimaryTHR = define_state(
    utility = 0,
    cost = 394
  ),
  SuccessP = define_state(
    utility = discount(u_SuccessP, .015),
    cost = 0
  ),
  RevisionTHR = define_state(
    utility = discount(u_RevisionTHR, .015),
    cost = discount(c_RevisionTHR, .06)
  ),
  SuccessR = define_state(
    utility = discount(u_SuccessR, .015),
    cost = 0
  ),
  Death = define_state(
    utility = 0,
    cost = 0
  )
)

mod_np1 <- define_model(
  transition_matrix = mat_np1,
  PrimaryTHR = define_state(
    utility = 0,
    cost = 579
  ),
  SuccessP = define_state(
    utility = discount(u_SuccessP, .015),
    cost = 0
  ),
  RevisionTHR = define_state(
    utility = discount(u_RevisionTHR, .015),
    cost = discount(c_RevisionTHR, .06)
  ),
  SuccessR = define_state(
    utility = discount(u_SuccessR, .015),
    cost = 0
  ),
  Death = define_state(
    utility = 0,
    cost = 0
  )
)

res_mod <- run_models(
  standard = mod_standard,
  np1 = mod_np1,
  parameters = param,
  cycles = 60,
  cost = cost,
  effect = utility,
  method = "end"
)
```

# Introduction

The goal of demographic analysis is to estimate the values of a model on an entire population. In `heemod` models are either run on a single set of parameters with `run_models()`, or on parameter distributions with `run_probabilistic()`. Demographic analysis runs the model on multiple sets of parameters (reflecting differents characteristics found in the target population) and combines the results.

# Population characteristics

The characteristics of the population are input from a table, with one column per parameter and one row per population strata. Here we define a simple population structure, with varying ages and sexes:

```{r}
pop_tab <- data.frame(
  age = rep(seq(30, 80, 10), 2),
  sex = rep(0:1, each = 6)
)
pop_tab
```

An optional value, `.weights`, allows to specify the respective weight of each strata in the target population:

```{r}
pop_tab$.weights <- c(
  3322, 5307, 7897, 1758,
  8878, 4052, 8555, 4120,
  4004, 5287, 9030, 8779
)
pop_tab
```

# Running model

For this example we will use the result from the assessment of a new total hip replacement previously described in `vignette("non-homogeneous", package = "heemod")`. `res_mod`, the result we obtained from `run_models()` in this previous vignette can be passed to `run_demographic()` to compute population-level results.

```{r message=FALSE}
res_mod

res_demo <- run_demographics(res_mod, pop_tab)
```

# Interpreting results

We can then print or plot the result of `run_demographics()` exactly like the results from `run_models()`.

```{r, fig.width = 6, fig.height=4, fig.align='center'}
res_demo

plot(res_demo, model = "np1")
```
