---
title: "Probabilistic Uncertainty Analysis"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Probabilistic Uncertainty Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, include=FALSE}
library(heemod)
```

This vignette shows how to transform the deterministic Markov model presented in `vignette("c-homogeneous", "heemod")` in a probabilistic model.

## Model definition

We will start by re-specifying the deterministic model of HIV therapy described previously (a monotherapy strategy `mono` and combined therapy strategy `comb`).

But instead of defining transition probabilities and state values directly in `define_transition()` or `define_state()` (as in the previous vignette), parameters will be defined first in a `define_parameters()` step. This is because only parameters defined this way can be resampled in a probabilistic analysis.

```{r}
param <- define_parameters(
  rr = .509,
  
  p_AA_mono = .721,
  p_AB_mono = .202,
  p_AC_mono = .067,
  p_AD_mono = .010,
  
  p_BC_mono = .407,
  p_BD_mono = .012,
  
  p_CD_mono = .250,
  
  
  p_AB_comb = p_AB_mono * rr,
  p_AC_comb = p_AC_mono * rr,
  p_AD_comb = p_AD_mono * rr,
  
  p_BC_comb = p_BC_mono * rr,
  p_BD_comb = p_BD_mono * rr,
  
  p_CD_comb = p_CD_mono * rr,
  
  p_AA_comb = 1 - (p_AB_comb + p_AC_comb + p_AD_comb),
  
  
  cost_zido = 2278,
  cost_lami = 2086,
  
  cost_A = 2756,
  cost_B = 3052,
  cost_C = 9007
)
```

We need to define `p_AA_mono` and `p_AA_comb` in `define_parameters()` because we will need to resample that value. Only values defined with `define_parameters()` can be resampled. So we cannot use the complement alias `C` to specify `p_AA_comb` in `define_transition()`, as we did before.

```{r}
mat_trans_mono <- define_transition(
  p_AA_mono, p_AB_mono, p_AC_mono, p_AD_mono,
  0,         C,         p_BC_mono, p_BD_mono,
  0,         0,         C,         p_CD_mono,
  0,         0,         0,         1
)
mat_trans_comb <- define_transition(
  p_AA_comb, p_AB_comb, p_AC_comb, p_AD_comb,
  0,         C,         p_BC_comb, p_BD_comb,
  0,         0,         C,         p_CD_comb,
  0,         0,         0,         1
)
```

State definition remains the same in this example.

```{r}
A_mono <- define_state(
  cost_health = cost_A,
  cost_drugs = cost_zido,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
B_mono <- define_state(
  cost_health = cost_B,
  cost_drugs = cost_zido,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
C_mono <- define_state(
  cost_health = cost_C,
  cost_drugs = cost_zido,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
D_mono <- define_state(
  cost_health = 0,
  cost_drugs = 0,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 0
)

A_comb <- define_state(
  cost_health = cost_A,
  cost_drugs = cost_zido + cost_lami,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
B_comb <- define_state(
  cost_health = cost_B,
  cost_drugs = cost_zido + cost_lami,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
C_comb <- define_state(
  cost_health = cost_C,
  cost_drugs = cost_zido + cost_lami,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 1
)
D_comb <- define_state(
  cost_health = 0,
  cost_drugs = 0,
  cost_total = discount(cost_health + cost_drugs, .06),
  life_year = 0
)
```

Strategies must be first defined and run as in a standard deterministic analysis.

```{r}
strat_mono <- define_strategy(
  transition = mat_trans_mono,
  A_mono,
  B_mono,
  C_mono,
  D_mono
)

strat_comb <- define_strategy(
  transition = mat_trans_comb,
  A_comb,
  B_comb,
  C_comb,
  D_comb
)

res_mod <- run_model(
  mono = strat_mono,
  comb = strat_comb,
  parameters = param,
  cycles = 50,
  cost = cost_total,
  effect = life_year
)
```

## Resampling distributions

Now we can define the resampling distributions. The following parameters will be resampled:

  * Relative risk.
  * Costs (such that cost are always positive).
  * Transition probability from AIDS to death.
  * The transition probabilities from state A.

Since the log of a relative risk follows a lognormal distribution, relative risk follows a lognormal distribution whose mean is `rr` and standard deviation on the log scale can be deduced from the relative risk confidence interval.

$$rr \sim lognormal(\mu = .509, \sigma = .173)$$

Programmed as:

```r
rr ~ lognormal(mean = .509, sdlog = .173)
```

Usually costs are resampled on a gamma distribution, which has the property of being always positive. Shape and scale parameters of the gamma distribution can be calculated from the mean and standard deviation desired in the distribution. Here we assume that *mean = variance*.

$$cost_A \sim \Gamma(\mu = 2756, \sigma = \sqrt{2756})$$

This can be programmed as:

```r
cost_A ~ make_gamma(mean = 2756, sd = sqrt(2756))
```

Proportions follow a binomial distribution that can be estimated by giving the mean proportion and the size of the sample used to estimate that proportion with `p_CD ~ prop(prob = .25, size = 40)`.

Finally multinomial distributions are declared with the number of individuals in each group in the sample used to estimate the proportions. These proportions follow a Dirichlet distribution:

  * `p_AA + p_AB + p_AC + p_AD ~ multinom(721, 202, 67, 10)`

```{r}
rsp <- define_psa(
  rr ~ lognormal(mean = .509, sdlog = .173),
  
  cost_A ~ make_gamma(mean = 2756, sd = sqrt(2756)),
  cost_B ~ make_gamma(mean = 3052, sd = sqrt(3052)),
  cost_C ~ make_gamma(mean = 9007, sd = sqrt(9007)),
  
  p_CD_base ~ prop(prob = .25, size = 40),
  
  p_AA_base + p_AB_base + p_AC_base + p_AD_base ~ multinom(721, 202, 67, 10)
)
```

## Run probabilistic model

Now that the distributions of parameters are set we can simply run the probabilistic model as follow:

```{r}
pm <- run_psa(
  model = res_mod,
  resample = rsp,
  N = 100
)
```

The average results are computed. In theory these values are more accurate than simple estimates because of non-linearities. An optional `threshold` can be passed to `summary()` to compute net monetary benefit.

```{r}
summary(pm, 
        threshold = c(1000, 5000, 6000, 1e4))
```

## Result interpretation

The results of the analysis can be plotted on the cost-effectiveness plane. We can see there seem to be little uncertainty on the costs compared to the uncertainty on the effects, resulting in an uncertainty cloud that looks like a line.

```{r, fig.width = 6, fig.height=4, fig.align='center'}
plot(pm, type = "ce")
```

And as cost-effectiveness acceptability curves or EVPI:

```{r, fig.width = 6, fig.align='center'}
plot(pm, type = "ac", max_wtp = 10000, log_scale = FALSE)
plot(pm, type = "evpi", max_wtp = 10000, log_scale = FALSE)
```

To compute EVPPI the results can also be exported with `export_savi()` in a format compatible with the [SAVI](http://savi.shef.ac.uk/SAVI/) software (Sheffield Accelerated Value of Information).

A covariance analysis can be performed:

```{r, fig.width = 6, fig.height = 4, fig.align='center'}
plot(pm, type = "cov")
```

As usual plots can be modified with the standard `ggplot2` syntax.

```{r, fig.align='center', fig.height=4, fig.width=6, message=FALSE}
library(ggplot2)

plot(pm, type = "ce") +
  xlab("Life-years gained") +
  ylab("Additional cost") +
  scale_color_brewer(
    name = "Strategy",
    palette = "Set1"
  ) +
  theme_minimal()
```

## Parallel computing

Resampling can be significantly sped up by using parallel computing. This can be done in the following way:

  * Define a cluster with the `use_cluster()` functions (i.e. `use_cluster(4)` to use 4 cores).
  * Run the analysis as usual.
  * To stop using parallel computing use the `close_cluster()` function.
  
Results may vary depending on the machine, but we found speed gains to be quite limited beyond 4 cores.
