---
title: "An Introduction to heemod"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{An Introduction to heemod}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, include=FALSE}
library(heemod)
```

This document is an introduction to `heemod`'s basic steps to define and run a model.

When building a Markov model for health economic evaluation, the following steps must be performed:

  * Specify transition probabilities between states.
  * Specify values attached to states (costs, utilities...).
  * Combine this information and run the model.
  
Other vignettes provide more details and examples on specific topics:

  * Simple Markov models: `vignette("homogeneous", package = "heemod")`: How to specify, run and interpret a simple Markov model with no time-varying elements.
  * Time-varying Markov models: `vignette("non-homogeneous", package = "heemod")`: Example analysis of a more complex Markov model with time-varying transition probabilities.

# Transition probabilities

The probability to change from one state to another during a time period is called a *transition probability*. The time period is called a *cycle*.

Transition probabilities between states can be specified through a 2-way table where the lines correspond to the states at the beginning of a cycle and the columns to the states at the end of a cycle. For example consider a model with 2 states `A` and `B`:

|       | A | B |
|-------|---|---|
| **A** | 1 | 2 |
| **B** | 3 | 4 |

When starting a cycle in state `A` (row `A`), the probability to still be in state `A` at the end of the cycle is found in colunm `A` (cell `1`) and the probability to change to state `B` is found in column `B` (cell `2`).

Similarly, when starting a cycle from state `B` (row `B`), the probability to be in state `A` or `B` at the end of the cycle are found in cells `3` or `4` respectively.

In the context of Markov models, this 2-way table is called a *transition matrix*. A transition matrix can be defined easily in `heemod` with the `define_matrix` function. If we consider the previous example, where cell values have been replaced by actual probabilities:

|       |  A  |  B  |
|-------|-----|-----|
| **A** | 0.9 | 0.1 |
| **B** | 0.2 | 0.8 |

That transition matrix can be defined with the following command:

```{r}
mat_trans <- define_matrix(
  .9, .1,
  .2, .8
)
mat_trans
```

# Attach values to states

In health economic evaluation, values are attached to states. Cost and utility are classical examples of such values. To continue with the previous example, the following values can be attachd to state `A` and `B`:

  * State `A` has a cost of 1234 per cycle and an utility of 0.85.
  * State `B` has a cost of 4321 per cycle and an utility of 0.50.
  
A state and its values can be defined with `define_state`:

```{r}
state_A <- define_state(
  cost = 1234,
  utility = 0.85
)
state_A

state_B <- define_state(
  cost = 4321,
  utility = 0.50
)
state_B
```

# Combine information in a model

Now that the transition matrix and the state values are defined, we can combine them to create a Markov model with `define_model`:

```{r}
mod_1 <- define_model(
  transition_matrix = mat_trans,
  state_A,
  state_B
)
mod_1
```

# Run a model

The model can then be run with `run_model` for a given number of cycles. The variables corresponding to valuation of cost and effect must be given at that point.

```{r}
res_mod_1 <- run_models(
  mod_1,
  cycles = 10,
  cost = cost,
  effect = utility
)
res_mod_1
```

By default the model is run for 1000 persons starting in the first state (here state `A`).

# Analyse results

We can plot the state membership counts over time. Other plot types are available.

```{r, fig.width = 6, fig.height=4, fig.align='center'}
plot(res_mod_1)
```

Plots can be modified using `ggplot2` syntax.

```{r, fig.align='center', fig.height=4, fig.width=6, message=FALSE}
library(ggplot2)

plot(res_mod_1) +
  xlab("Time") +
  ylab("N") +
  theme_minimal() +
  scale_color_brewer(
    name = "State",
    palette = "Set1"
  )
```

# Convenience functions

Convenience functions are avaible to make it easier to compute transition probabilities from indidence rates, OR, RR, or probabilities estiamted on a different timeframe.

Example : convert an incidence rate of 162 cases per 1,000 person-years to a 5-year probability.

```{r}
rate_to_prob(r = 162, per = 1000, to = 5)
```

Type `?probability` to see a list of the convenience functions available.

# Going futher

In order to compare different strategies it is possible to run several models in parallel, examples are provided in `vignette("homogeneous", package = "heemod")` or `vignette("non-homogeneous", package = "heemod")`.

Uncertainty analysis `vignette("probabilistic", package = "heemod")` and sensitivity analysis `vignette("sensitivity", package = "heemod")` can be performed.

See vignette `vignette("reproduction", package = "heemod")` for an exact reproduction of the analyses from Decision Modelling for Health Economic Evaluation.

Population-level country-specific mortality rates by age and sex (often used as transition probabilities in Markov models) can be downloaded from WHO databases with the `get_who_mr()` function.
