---
title: "Running survival models from tabular data"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true 
vignette: >
  %\VignetteIndexEntry{Survival models}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
library(heemod)
knitr::opts_chunk$set(echo = TRUE)
library(flexsurv)
```

Running an oncology cost/benefit model in heemod involves several steps:

* Fit available survival data under different distributional assumptions.

* Examine the survival fits and decide which ones are most appropriate to use in the economic model.

* Run the economic model in heemod with the chosen survival fits.

This document will demonstrate how to accomplish those steps in heemod,
using simulated data sets included with the package.

```{r fits}
base_dir <- system.file("tabular\\surv", package = "heemod")
spec_file <- "example_oncSpecs_nomatrix.csv"
```

Other vignettes cover more general information on the files needed to run heemod from [tabular input](h-tabular.html), and additional
information on [survival curve fitting](j-survival.html). 
Here we will deal only with those aspects specific to fitting survival curves
from data and using them to run economic models.


## Specifying the data to fit

The source of the survival data, as well as the names and locations of the resulting fits, 
are specified in a file with one row for each (treatment, outcome) pair, where the outcome can be
either progression-free survival (PFS) or overall survival (OS).
 Here’s an example of such a file, pointing to an example using randomly-generated data
 and two fake treatments, “A” and “B”:

```{r surv-info}
surv_inf_file_name <- "survival_info.csv"
full_path <- file.path(base_dir, surv_inf_file_name)
surv_info <- heemod:::read_file(full_path)
print(surv_info)
```

The `data_directory` and `data_file` columns allow us to specify the location
of data separately for each row.    The `fit_directory`, `fit_file`, and 
`fit_name` columns, similarly, allow us specify where the fits should be kept,
and what the R object containing the fits should be called.   Finally, `time_col`,
`treatment_col`, and `censor_col` give us the names of the columns in the data
files that contain the three elements needed to specify the model.  Together,
these give us a great deal of flexibility.   For example, we may keep our data
in a set of directories that are, for most analysts, read-only, while still
allowing them to save their fits (each analyst can even save their own fits,
though it seems preferable to have them shared).   We can have data for
different treatments and outcomes in different files or in the same file -
if they are in the same file, we can specify the columns to be used
to specify the different outcomes.

## Specifying multiple subsets of data for survival models

Some analyses require fitting survival curves for different subsets of the
data.  These subsets may be based on demographic or potential biomarker
variables - anything that could potentially influence survival.  The
file defining the subsets should be in the fit directory where the fits
will be stored (which may often be the same place where the raw data is
stored).  That allows different people to fit different subsets
if necessary, and also allows for defining subsets if the data directory
is read-only for the analyst.  Currently the file must be called 
"set_definitions.X", where "X" can be ".csv", ".xls", or ".xlsx" depending
on the file format.

Below is an example:
```{r subset-file}
subset_def_file_name <- "set_definitions.csv"
full_path <- file.path(base_dir, "survival_data", subset_def_file_name)
subset_info <- heemod:::read_file(full_path)
print(subset_info)
```
The file must contain three columns: treatment, set_name, and condition.
"treatment" must match the treatment column in the survival information
file, because that is how fits will be associated to the appropriate
treatments.  The set_name and condition files define the names of the
subsets, which will be used later in specifying which fits to use in
a particular analysis, and the condition defining the subset.  The condition
must be a statement that can be parsed in R, and variables used should
match the relevant column names in the survival data file.   For example,
the file presented here uses "time" in rows 3 and 4 because that is the column
name in our
example data files.  If the example data used "Time" or "t" or "event_time", 
the condition statements would need to use those names instead.

One notable case highlighted here is subsets based on time itself - for example,
we may wish to fit the a survival model only for event times after some chosen time.
In this case, we will generally want to fit the survival model as if the
chosen time were time 0.   We can do this by adding the column `time_subract`
with the appropriate number.   This currently requires the analyst to take
care that the name of the subset, the condition defining the subset, and
the time subtracted all match up.  For example, we would not want to
name a subset `time.gt.50` but then define it using the condition `time > 100`,
nor would we want to assign a `time_subtract` other than 50 to a subset with
that name.   In the future, we may automatically create the time condition
based on `time_subract`, but names will still need to be checked by the
analyst.   If `time_subtract` is not provided in the file, or is missing
in any given row, if it assumed to be 0.

If the `set_definition` file is missing, then a single subset, called "all",
with condition equal to `TRUE` (which takes all points) will be fit 
for each treatment.   If the `set_definition` file is present, it should
include the "all" subset, as in the first two rows of the example above.

The example also demonstrates that we need not define the same subsets,
or even the same number of subsets, for each treatment.  Here, although
both A and B have values for "biomarker" in the files, we choose to 
define a subset based on "biomarker" only for treatment B.  
All the fits for all the defined subsets will be fit at the same time,
and stored in a data frame (actually a tibble, for those interested
in the technical details) for future use.

## Fitting survival models

The function `partitioned_survival_from_tabular` fits the survival models 
to data.
Here we use `save_fits = TRUE`; it is also possible to set `save_fits = FALSE`
to obtain the fits in R without saving them to a file. The specification file
will be used again to run the economic model, but to create the fits it needs
only the `tm` field. `tm` stands for "transition matrix", because the `heemod` package
initially took only Markov transition matrices; now it can take the information
to specify partitioned survival models also - in this case, the survival information
file just discussed.  If desired, the rest of the fields can be
added later.

The distributions currently used for survival fits are exponential, Weibull, 
lognormal, log-logistic, gamma, Gompertz, and generalized gamma. 
If fitting fails for some distribution, `heemod` will note the error and 
continue, so results will be available for the other distributional assumptions.
(Here we set `save_fits = FALSE` because we've already saved the fits in
the package.)

```{r make-fits}
temp1 <- survival_fits_from_tabular(base_dir, spec_file, 
                                    save_fits = FALSE)[[1]]
```

`heemod` does not automatically choose a distribution to use in the survival model - the modeler must choose the most appropriate distribution for each set of survival data.
Fit quality can be examined using the negative log-likelihood, the Aikake
and Bayes information critera, and various kinds of graphs. Below we remove the `fit`, `set_def`, and `time_subract` columns so the fit metrics are visible.   Note that fit metric values are not
generally comparable across different subsets, which are fit on different subsets of data.
They are more appropriate for comparing across different models fit to the same data.

```{r}
names(temp1)
```

```{r}
extract_surv_fit_metrics(temp1, c("AIC", "BIC" ,"m2LL"))[, -c(5,6, 7)]
```

## Specifying which fits to use in a given analysis

When a modeler has decided to run the economic analysis using a particular set of fits, the specification must be included in a file.  The name of the file goes in the
specification file, in a row with `data` element `use_fits`.

The file has three or four columns: \itemize{
\item{`.strategy`, corresponding to the strategies in the model;}
\item{`.type`, which for each row is either 'pfs' or 'os';}
\item{`dist`,  specifying the survival distribution to use;}
\item{`until` (optionally), specifying when the distribution ceases to apply.}
}
`until` needs to be specified only if there is more than one row with the same
`.strategy` and `.type`.  In that case, each distribution will pick up where
the previous one left off.   (The rows need not be presented in order, though
that is a good practice for readability; the code will sort distributions into
order by `until`.)   Here we don't require it.

If we define `use_fits` as
```{r define_fits}
use_fits_file <- file.path(base_dir, "use_fits_example0.csv")
use_fits <- heemod:::read_file(use_fits_file)
use_fits
```

then the analysis will use an exponential survival model for overall survival for both strategies, but will use the gamma survival model for progression-free survival for A, and a lognormal survival model for progression-free survival for B.

We can also use `fit('km')` to get the Kaplan-Meier curve for a given fit.

## Running the analysis using already-fit survival models

The same specification file that was used for `survival_fits_from_tabular` can be used for `run_model_tabular`. `run_model_tabular` uses the information in the specification file to pick up the desired fits from those created earlier. No survival fitting happens when running  run_model_tabular, only selection.  The selection is governed by the `use_fits` element of the reference file.

Here is the use_fits element of this file:
```{r}
ref <- heemod:::read_file(file.path(base_dir, spec_file))
use_fits_file <- file.path(base_dir, ref[ref$data == "use_fits", "file"])
use_fits <- heemod:::read_file(use_fits_file)
use_fits
```

```{r}
surv_example <- run_model_tabular(base_dir, spec_file,
                                  run_psa = FALSE, run_demo = FALSE,
                                  save = FALSE, overwrite= FALSE)
```
The model has run as requested:
```{r fig.width=8, fig.height=5}
plot(surv_example$model_runs)
```

We can also mix explicit survival distribution definitions with fits,
as shown below.
```{r}
spec_file <- "example_oncSpecs_nomatrix_mixed.csv"
ref <- heemod:::read_file(file.path(base_dir, spec_file))
mixed_use_fits_file <- file.path(base_dir, ref[ref$data == "use_fits", "file"])
mixed_use_fits <- heemod:::read_file(mixed_use_fits_file)
mixed_use_fits
```

```{r fig.width = 8, fig.height = 5, run_mixed_dist}
surv_example_mixed <- run_model_tabular(base_dir, spec_file,
                                  run_psa = FALSE, run_demo = FALSE,
                                  save = FALSE, overwrite= FALSE)
plot(surv_example_mixed$model_runs)
```
We can also specify more complex models, using all of the modifications to
survival fits that are described in [the survival vignette](j-survival.html).
In particular, we can use the Kaplan-Meier curve until a particular time
and then switch over to a fit, as shown below.

```{r}
spec_file <- "example_oncSpecs_nomatrix_join.csv"
ref <- heemod:::read_file(file.path(base_dir, spec_file))
join_use_fits_file <- file.path(base_dir, ref[ref$data == "use_fits", "file"])
join_use_fits <- heemod:::read_file(join_use_fits_file)
join_use_fits
```
In this case, we see the less regular changes of the Kaplan-Meier curve until
time 50, followed by the more regular changes of the fit model.
```{r fig.width = 8, fig.height = 5, run_joined_dist}
surv_example_mixed <- run_model_tabular(base_dir, spec_file,
                                  run_psa = FALSE, run_demo = FALSE,
                                  save = FALSE, overwrite= FALSE)
plot(surv_example_mixed$model_runs)
```
