---
title: "Time-varying Markov models (non-homogenous)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time-varying Markov models (non-homogenous)}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE, include=FALSE}
library(heemod)
```

With more complex Markov models in health economic evaluation, transition probabilities between states can vary with time. Those are called *non-homogenous* or *time-inhomogenous* Markov models. A further distinction can be made:

  1. transition probabilities depend on how long __the entire model__ has been running (model-time varying): this situation can be modelled with a non-homogenous Markov model;
  2. transition probabilities depend on how long __an individual__ has been in a state (individual-time varying): this situation usually needs to be modelled with a microsimulation.

However in some special cases non-homogenous Markov models can be used in situation 2. This is when the state having individual-time varying transition probabilities is the starting state __and__ it is not possible to go back to that state after having left it. In this situation individual-time is equivalent to model-time. This is the case for this example.

If you are not familiar with `heemod`, first consult the [introduction vignette](https://cran.r-project.org/web/packages/heemod/vignettes/introduction.html).

# Model description

This example is an implementation of the assessment of a new total hip replacement (THR) technology described in chapter 3.5 of [Decision Modelling for Health Economic Evaluation](http://ukcatalogue.oup.com/product/9780198526629.do). A more detailed report is available [at this location](https://www.york.ac.uk/media/che/documents/papers/technicalpapers/CHE%20Technical%20Paper%2028.pdf), event though this reports goes a bit further in the analysis.

This model has 5 states:

  * Primary THR: starting state, individuals receive either the standard or the new THR, outcomes are either succes of primary THR or death (operative mortality rate of primary THR `omrPTHR`);
  * Success of primary THR: state after receiving primary THR if the surgery is successfull, individuals can stay in this state, need a THR revision, or die;
  * Revision of primary THR: for individuals whose primary THR failed, outcomes are either success of revision THR or death (operative mortality rate of revision THR `omrRTHR`);
  * Success of revision THR: state after receiving revision THR if the surgery is successfull, individuals can stay in this state, need a THR re-revision, or die;
  * Death, either caused by THR or another cause.
  
Two transition probabilities are time-varying in this model:

  * Probability of death by another cause increases with age;
  * Probability of primary THR revision increases with time.
  
Other-cause death probabilities (mortality rate `mr`) vary with age and gender with the following values:

```{r}
death_prob <- data.frame(
  age = rep(seq(35, 85, 10), each = 2),
  sex = rep(0:1, 6),
  mr = c(
    1.51e-3, .99e-3, 3.93e-3,
    2.6e-3, 10.9e-3, 6.7e-3,
    31.6e-3, 19.3e-3, 80.1e-3,
    53.5e-3, 187.9e-3, 154.8e-3
  )
)
death_prob
```

Primary THR revision probability increases with time with the following formula (a Weibull distribution):

$$
P_{revision} = 1 - \exp(\lambda \times ((t-1)^\gamma-t^\gamma))
$$

Where $t$ is the time since revision, $\gamma = 1.45367786$ and:

$$
\lambda = exp(cons + ageC \times age + maleC \times sex)
$$

Where `age` and `sex` (female = 0, male = 1) are individual characteristics, `cons` = -5.49094, `ageC` = -0.0367 and `maleC` = 0.768536.

Revision THR re-revision (`rrr`) probability is set to be constant at 0.04 per year.

# Defining parameters

The key element to specify time-varying elements in `heemod` is through the use of a package-defined variable, `markov_model`. This variable takes increasing values with each cycles, starting from 1. For example the age of individuals at any moment can be defined as `Initial age + markov_cycle`.

In order to build this more complex Markov model, parameters need to be defined through `define_parameters`. The equations decribed in the previous section can be simply written, here for a female population (`sex` = 0) starting at 60 years old (`age_init` = 60):

```{r}
# a function to return mr, given age and sex
mr_func <- function(age, sex)
  merge(data.frame(age = floor(age/10-.5)*10+5, sex = sex), death_prob)$mr

param_thr <- define_parameters(
    t = markov_cycle,
    
    age_init = 60,
    sex = 0,
    age = age_init + t,
    
    omrPTHR = .02,
    omrRTHR = .02,
    rrr = .04,
    
    cons = -5.49094,
    ageC = -.0367,
    maleC = .768536,
    lambda = exp(cons+ageC*age_init+maleC*sex),
    gamma = 1.45367786,
    
    # rrNP1 = .260677,
    
    standardRR = 1 - exp(lambda*((t-1)^gamma-t^gamma)),
    
    mr = mr_func(age_cat, sex)
)
param_thr
```


